---
title: "Data Wrangling"
author: "by [Yan Holtz](https://github.com/holtzy/) - `r format(Sys.time(), '%d %B %Y')`"
---

  
```{r echo=FALSE}
# Just to add space between the introduction of the document
knitr::asis_output(htmltools::htmlPreserve("<br><br>"))
```


This file is the first part of the UK Biobank comorbidity analysis. It does all the boring data wrangling part of the Job:  

* Create a Phenotype file from the whole UKB (**server only**)
* Take the ICD10 classification (see [here](https://holtzy.github.io/Visualizing-the-ICD10-Classification/))
* Explain how HES data were harvested and treated  
* Calculate the Comorbidty matrix: gives the occurence of each **disease** and each **pair of disease** for the 3 levels of classification (data available [here](https://github.com/holtzy/UKB-Comorbidity/tree/master/0_DATA))  
* Get heritability estimates of 303 diseaes from Neale's Lab (data available [here](https://github.com/holtzy/UKB-Comorbidity/tree/master/0_DATA))


```{r, message=FALSE, warning=FALSE}
# We will need a few library to run this document
library(tidyverse)
library(rmarkdown)    # You need this library to run this template.
library(epuRate)      # Install with devtools: install_github("holtzy/epuRate", force=TRUE)
library(xtable)
library(knitr)
```








#Harvest the UKB data
***
The first step takes place in the QBI cluster. I must harvest the fields I need in the whole UKB phenotype data set. To do so I read the file 9280_1250_UKBiobank.html. I select the field that interest me. Then I harvest this field using this Bash code. Note that I have to add 1 to the 'Column' column that I select.


First harvest in August 2017:
```{r eval=FALSE}
# Version 1: August 2017
zcat /references/UKBiobank/pheno/download/9280_12505_UKBiobank.tab.gz | cut -f1,8,9,22-24,144-146,147-149,4069-4071,4081,4082,4066-4068,3438,3439-3441,5899-5901,5902-5943,6181,6185-6564,6593-7027,7058-
7062,26-28,29-31,4075-4077,4081,5896-5898,6008-6010,6179,6180,262-264,361-363 > /ibscratch/wrayvisscher/Yan/UKB_phenos.txt
```

Second harvest in October: I add the ICD10 secondary diagnosis
```{r eval=FALSE}
# Version 2: October 2017

# Make script
cd /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS
nano extract.sh
zcat /references/UKBiobank/pheno/download/9280_12505_UKBiobank.tab.gz | cut -f1,8,9,22-24,144-146,147-149,4069-4071,4081,4082,4066-4068,3438,3439-3441,5899-5901,5902-5943,6181,6185-6564,6593-7027,7058-7062,26-28,29-31,4075-4077,4081,5896-5898,6008-6010,6179,6180,262-264,361-363,6593-7027 > /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/UKB_phenos.txt
# Send with Qsub
qsub -b y -cwd -e tmp -o tmp "/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS/extract.sh"
# Zip the result
gzip /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/UKB_phenos.txt
```

Files are transfered locally by scp:
```{r eval=FALSE}
# Transfert locally
cd /Users/y.holtz/Dropbox/QBI/3_UK_BIOBANK_COMO_PROJECT/1_FILES
scp  uqyholtz@cluster.qbi.uq.edu.au://ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/UKB_phenos.txt.gz .
# Note: file = 39G
```







#ICD 10 Classification 
***
The human diseases are classified by the [WHO](https://en.wikipedia.org/wiki/World_Health_Organization) (World Health Organization) disease classification ( International Statistical Classification of Diseases: [ICD](http://apps.who.int/classifications/icd10/browse/2010/en)). About 16k diseases are reported and coded.  

This classification has been downloaded, cleaned and describe in [this document](https://holtzy.github.io/Visualizing-the-ICD10-Classification/).






#Hospital Episode Data (HES)
***
Suppose somebody has 3 ICD10 diseases. We are highly interested to know WHEN each disease was discovered. The phenotype file provides us with the number of time somebody has been to the hospital. But an external file is needed to know when it was.  
The HES (Hospital Episode Statistics) data provides this information.  

Load it:
```{r}
#HES=read_tsv("/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/RECORD_REPOSITORY/ukb_hesin.tsv")
```

Note that we need a bridge file to make the links with the phenotype because this info comes from the application of John, not the one from Naomi and Peter like for the phenotype.
```{r}
#bridge=read_delim("/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/ukb_bridgingfile_12505_10214.txt", delim=" ")
```











#Comorbidity Matrix
***


##Aggregate diseases
We want to study comorbidity at different levels: group, subgroup and subsubgroup. The idea is thus to read the UKB phenotypes and replace each disease ID (currently subsubgroup) per its subgroup and group id.
This is done through a custom Python script. I make this job 4 times: subsubgroups, subgroup, group, and mental vs other.

The first step is to prepare to "equivalence" files: what do I replace by what?
```{r, eval=FALSE}
load("/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/WHO_disease_classification_clean.R")
library(dplyr)

# L3 (subgroup). In the UKB I can find codes at levels 4 (mainly) OR at level 5 (sometimes). Thus I must take both into account
# When I have code at level3 I can just keep them!
a=ICD[ , c("coding_L4", "coding_L3" )]
b=ICD[ , c("coding_L5", "coding_L3" )] 
colnames(b)=colnames(a)=c("from", "to")
tmp=na.omit(rbind(a,b))
write.table(tmp, "/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/equivalence_L3.txt", quote=F, row.names=F, sep="\t")

# L2 (group)
a=ICD[ , c("coding_L3", "coding_L2" )]
b=ICD[ , c("coding_L4", "coding_L2" )]
c=ICD[ , c("coding_L5", "coding_L2" )] 
colnames(c)=colnames(b)=colnames(a)=c("from", "to")
tmp=na.omit(rbind(a,b,c))
write.table(tmp, "/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/equivalence_L2.txt", quote=F, row.names=F, sep="\t")

# L1 (mental, others..)
a=ICD[ , c("coding_L3", "short" )]
b=ICD[ , c("coding_L4", "short" )]
c=ICD[ , c("coding_L5", "short" )] 
colnames(c)=colnames(b)=colnames(a)=c("from", "to")
tmp=na.omit(rbind(a,b,c)) %>% unique()
write.table(tmp, "/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/equivalence_L1.txt", quote=F, row.names=F, sep="\t")
```

Now we can make the transformation with the python script
```{r eval=FALSE}
cd /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX

# level 3
qsub -b y -cwd -N tmp_L3 "/clusterdata/apps/python2.7.1/Python-2.7.1/python  /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS/script_replace_values.py -input /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/UKB_phenos.txt.gz -replace /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/equivalence_L3.txt -output /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/UKB_phenos_L3.txt"

# level 2
qsub -b y -cwd -N tmp_L2 "/clusterdata/apps/python2.7.1/Python-2.7.1/python  /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS/script_replace_values.py -input /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/UKB_phenos.txt.gz -replace /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/equivalence_L2.txt -output /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/UKB_phenos_L2.txt"

# level 1
qsub -b y -cwd -N tmp_L1 "/clusterdata/apps/python2.7.1/Python-2.7.1/python  /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS/script_replace_values.py -input /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/DATA/UKB_phenos.txt.gz -replace /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/equivalence_L1.txt -output /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/UKB_phenos_L1.txt"

# An compress
gzip UKB_phenos*
```







##Calculate occurence of pairs
Some individuals have several mental diseases. Let's check which disease tend to occur together? 
For each pair of mental disease, I'm going to count the number of time the pair appear.  
I build a square matrix. Each cell gives the number of time 2 diseases have been seen together. Not that easy to compute actually!  
This is a script that I sent on qsub on the cluster:  

```{r, eval=FALSE}

#!/usr/bin/env Rscript
# This is an Rscript that takes the UKB phenotype matrix as input
args = commandArgs(trailingOnly=TRUE)
input_file=args[1]
output_file=args[2]

# libraries
library(dplyr, lib.loc="/clusterdata/uqyholtz/R/x86_64-unknown-linux-gnu-library/3.1/")

# Read the data already takes minutes.
# Don't forget to change the number of lines
data=read.table(input_file, header=T, sep="\t")
print("File has been read. Number of row:")
print(nrow(data))

# List of every disease found (not only mental)
all=data %>% select(f.41202.0.0:f.41202.0.379  ,  f.41204.0.0:f.41204.0.434)  %>% as.matrix %>% as.vector() %>% unique() 
all=all[!is.na(all)]
print("number of diseases found:")
print(length(all))

# What's the time
start=Sys.time()

# Initialize a 0 matrix with all the pair
selection=data %>% select(f.41202.0.0:f.41202.0.379  ,  f.41204.0.0:f.41204.0.434)
output=matrix(0,length(all), length(all))
colnames(output)=all
rownames(output)=all

# What's the time
start=Sys.time()

# Apply this function to each line of the matrix
for(i in seq(1, nrow(data))){
  print(i)
  a=data[i,] %>% as.matrix %>% as.vector() %>% unique() 
  b=match(a, colnames(output))
  b=b[!is.na(b)]
  if(length(b)!=0){
    output[b,b]=output[b,b]+1
  }
}

# Write the result
write.table(output, file=output_file, sep=";", quote=F, col.names = T, row.names = F )

# What's the time. How long did it take?
end=Sys.time()
end-start
```

Send this script in qsub on the QBI cluster:
```{r eval=FALSE}
cd /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX

# L3
qsub -b y -cwd -e tmp_calc_como_adjacency_UKB_L3_log -o tmp_calc_como_adjacency_UKB_L3_log -l vf=30G,h_vmem=30G "/clusterdata/apps/R-3.2.3/bin/Rscript /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS/script_COMO_adjacency.R /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/UKB_phenos_L3.txt.gz /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/Como_Ocurrence_L3.txt"

# L2
qsub -b y -cwd -e tmp_calc_como_adjacency_UKB_L2_log -o tmp_calc_como_adjacency_UKB_L2_log -l vf=30G,h_vmem=30G "/clusterdata/apps/R-3.2.3/bin/Rscript /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS/script_COMO_adjacency.R /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/UKB_phenos_L2.txt.gz /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/Como_Ocurrence_L2.txt"

# L1
qsub -b y -cwd -e tmp_calc_como_adjacency_UKB_L1_log -o tmp_calc_como_adjacency_UKB_L1_log -l vf=30G,h_vmem=30G "/clusterdata/apps/R-3.2.3/bin/Rscript /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/SCRIPTS/script_COMO_adjacency.R /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/UKB_phenos_L1.txt.gz /ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/Como_Ocurrence_L1.txt"

#gzip the results
gzip Como*
```

Once the job is over, I take the result back locally for further analysis.
```{r eval=FALSE}
cd /Users/y.holtz/Dropbox/QBI/3_UK_BIOBANK_COMO_PROJECT/1_FILES
scp uqyholtz@cluster.qbi.uq.edu.au:/ibscratch/wrayvisscher/Yan_Holtz/3_UKB_COMO/COMORBIDITY_INDEX/Como_Ocurrence* .
```

Then, this matrix can be read as follow:
Read the matrix in R
```{r}
data=read.table("0_DATA/Como_Ocurrence_L1.txt.gz", sep=";", header=T)
data=as.matrix(data)
rownames(data)=colnames(data)
as.data.frame( data[1:5, 1:5] )
```

Three version of this count table are thus provided and available on [github](https://github.com/holtzy/UKB-Comorbidity/tree/master/0_DATA): at level1, level2 and level3. 







#Heritability
***
Heritabilities of 303 diseaseas have been recovered from the Neale's lab website.
We have 14 mental diseases here.
```{r}
Herit=read.table("0_DATA/ukbb_all_h2part_results.txt.gz", sep="\t", header = T)
Herit=Herit[ substr(Herit$phenotype, 1, 1) %in% LETTERS , c(1:20) ]
Herit=Herit %>% arrange(phenotype)
#dim(Herit)
#colnames(Herit)
Herit %>% head()
```











