---
title: "A clean and uncluttered template"
author: "by [Yan Holtz](https://github.com/holtzy/) - `r format(Sys.time(), '%d %B %Y')`"
output:
  epuRate::epurate:
    toc: FALSE
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
```{r echo=FALSE}
# Just to add space between the introduction of the document
knitr::asis_output(htmltools::htmlPreserve("<br><br>"))
```


> I find it not too bad to start with a couple of sentences of introduction. Explain briefly what this document is going to talk about. May be add a [useful link](www.r-graph-gallery.com) relative to this project


# Get the data
***
Usually I always start by loading libraries. The tidyverse is probably the package I use the most. Quite difficult to imagine analysing data without it. It is a good practice to load all the packages in the same time. It allows to know exactly what is needed to reproduce your analysis.
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(rmarkdown)    # You need this library to run this template.
library(epuRate)      # Install with devtools: install_github("holtzy/epuRate", force=TRUE)
```

Then I explain what is my input data and I load it. Here is how it looks like:
```{r}
# Show it:
mtcars %>% head(5)
```

This file has `r nrow(mtcars)` lines and `r ncol(mtcars)` columns. It is ready to be analysed.





# A few analysis {.tabset .tabset-fade .tabset-pills}
***
Here are 3 version of a circular barplot that you can find in the R graph gallery. Note that the figure is centered. You can change the size of the plot with **fig.width** and **fig.height**  

